\textbf{Effective application of machine learning to microbiome-based classification problems}

### 2019-10-21 submission/manuscript.tex
last edits before submission

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without a separate held-out test data; unreported variation
between the predictive performance on different folds of
cross-validation; and unreported variation between cross-validation and
testing performances. Though the microbiome field is making progress to
avoid some of these pitfalls including validating their models on
independent datasets (8, 19, 20) and introducing ways to better use ML
tools (21--24), more work is needed to improve reproducibility further
and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modeling approach has often been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
non-linear models such as random forest and deep neural networks (3, 12,
25--27) over simpler models such as logistic regression or other linear
models (19, 23, 28). Although in some cases, complex models may capture
important non-linear relationships and therefore yield better
predictions, they can also result in black boxes that lack
interpretability. Such models require post hoc explanations to quantify
the importance of each feature in making predictions. Depending on the
goal of the modeling, other approaches may be more appropriate. For
example, researchers trying to identify the microbiota associated with
disease may desire a more interpretable model, whereas clinicians may
(but not always) emphasize predictive performance. Nonetheless, it is
essential to understand that the benefit of more complex, less
interpretable models may be minimal (29, 30). It is important for
researchers to justify their choice of modeling approach.

@@

\textbf{Model selection and pipeline construction}. We established a
reusable ML pipeline for model selection and evaluation, focusing on
seven different commonly used supervised learning algorithms {[}Figure
1{]}.

@@

Model selection requires tuning hyperparameters. Hyperparameters are
parameters that need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that the model's predictions are no
different than random. To select hyperparameters, we performed a grid
search for hyperparameter settings when training the models. Default
hyperparameter settings in previously developed ML packages available in
R, Python, and MATLAB programming languages may be inadequate for
effective application of classification algorithms and should to be
optimized for each new ML task. For example, L1-regularized SVM with
linear kernel showed large variability between different regularization
strengths (C) and benefited from tuning {[}Figure S1{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
The predictive performance of random forest model was higher than other
ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
regression, XGBoost, L2-regularized SVM with linear and radial basis
function kernel AUROC values were not significantly different from one
another and had median AUROC values of 0.680 {[}IQR 0.639-0.750{]},
0.679 {[}IQR 0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668
{[}IQR 0.639-0.750{]}, respectively. L1-regularized SVM with linear
kernel and decision tree had significantly lower AUROC values than the
other ML models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and
0.601 {[}IQR 0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly,
these results demonstrate that the most complex model (XGBoost) did not
have the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well non-linear models.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
could indicate that the models were overfit to the training data. The
largest difference in median AUROCs was 0.021 in L1-regularized SVM with
linear kernel, followed by SVM with radial basis function kernel and
decision tree with a difference of 0.007 and 0.006, respectively
{[}Figure 2{]}. These differences are relatively small and give us
confidence in our estimate of the generalization performance of the
models.

@@

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. We did not take advantage of
parallelization in training, nor was the code optimized for speed.
Still, as expected, the training times increased with the complexity of
the model and the number of potential hyperparameter combinations. Also,
the linear models trained faster than non-linear models {[}Figures
S1-S2; Figure 5{]}.

@@

We established a pipeline that can be generalized to any modeling method
that predicts a binary health outcome. We performed a random data-split
to create a training set (80\% of the data) and a held-out test set
(20\% of the data), which we used to evaluate predictive performance. We
used the AUROC metric to evaluate predictive performance as our dataset
was highly balanced. We repeated this data-split 100 times to measure
the possible variation in predictive performance. During training, we
tuned the model hyperparameters with a repeated five-fold
cross-validation. Despite the high number of features microbiome
datasets typically have, the models we built with this pipeline
generalized to the held-out test sets.

@@

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
generated by Sze et al (44), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (42, 45, 46);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter. A grid
search was needed to avoid large variation in prediction performance.
For L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (47).

The computational burden during model training due to model complexity
was reduced by parallelizing segments of the ML pipeline. We
parallelized the training of each data-split. This allowed the 100
data-splits to be processed through the ML pipeline simultaneously at
the same time for each model. It is possible to further parallelize the
cross-validation step for each hyperparameter setting.

@@

\textbf{Acknowledgements.} We thank all the study participants of Great
Lakes-New England Early Detection Research Network. We would like to
thank the members of the Schloss lab for their valuable feedback. Salary
support for M.R came from NIH grant 1R01CA215574. Salary support for
P.D.S. came from NIH grants P30DK034933 and 1R01CA215574.

@@

\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
absolute feature weights of (A) L1-regularized SVM with linear kernel,
(B) L2-regularized SVM with linear kernel, and (C) L2-regularized
logistic regression, were ranked from highest rank, 1, to lowest rank,
100, for each data-split. The feature ranks of the 20 highest ranked
OTUs based on their median ranks (median shown in black) are reported
here. OTUs that were associated with classifying a subject as being
healthy had negative signs and were shown in blue. OTUs that were
associated with classifying a subject having an SRN had positive signs
and were shown in red. Some of the same OTUs were identified as
important in all of the linear models. \newpage
\includegraphics{Figure_4.png}

\textbf{Figure 5. Training times of seven ML models.} The median
training time was the highest for XGBoost and shortest for
L2-regularized logistic regression. \newpage
\includegraphics{Figure_S1.png}

\textbf{Figure S1. Hyperparameter setting performances for linear
models.} (A) L2-regularized logistic regression, (B) L1-regularized SVM
with linear kernel, and (C) L2-regularized SVM with linear kernel mean
cross-validation AUROC values when different hyperparameters were used
in training the model. The stars represent the highest performing
hyperparameter setting for each model. \newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}

\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices.}
\small

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied to.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline

@@

>>>>

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without separate held-out test data; unreported variation between
the predictive performance on different folds of cross-validation; and
unreported variation between cross-validation and testing performances.
Though the microbiome field is making progress to avoid some of these
pitfalls including validating their models on independent datasets (8,
19, 20) and introducing ways to better use ML tools (21--24), more work
is needed to improve reproducibility further and minimize overestimating
for model performance.

Among microbiome researchers, the lack of justification when selecting a
modeling approach has often been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
non-linear models such as random forest and deep neural networks (3, 12,
25--27) over simpler models such as logistic regression or other linear
models (19, 23, 28). Although in some cases, complex models may capture
important non-linear relationships and therefore yield better
predictions, they can also result in black boxes that lack
interpretability. Such models require post hoc explanations to quantify
the importance of each feature in making predictions. Depending on the
goal of the modeling, other approaches may be more appropriate. For
example, researchers trying to identify the microbiota associated with
disease may desire a more interpretable model, whereas clinicians may
emphasize predictive performance. Nonetheless, it is essential to
understand that the benefit of more complex, less interpretable models
may be minimal (29, 30). It is important for researchers to justify
their choice of modeling approach.

@@

\textbf{Model selection and pipeline construction}. We established a
reusable ML pipeline for model selection and evaluation, focusing on
seven different commonly used supervised learning algorithms {[}Figure
1, Table 1{]}.

@@

Model selection requires tuning hyperparameters. Hyperparameters are
parameters that need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that the model's predictions are no
different than random. To select hyperparameters, we performed a grid
search for hyperparameter settings when training the models. Default
hyperparameter settings in developed ML packages available in R, Python,
and MATLAB programming languages may be inadequate for effective
application of classification algorithms and need to be optimized for
each new ML task. For example, L1-regularized SVM with linear kernel
showed large variability between different regularization strengths (C)
and benefited from tuning as the default C parameter was 1 {[}Figure
S1{]}.

@@

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having healthy colons or SRNs {[}Figure 2{]}.
The predictive performance of random forest model was higher than other
ML models with a median 0.695 {[}IQR 0.650-0.739{]}, though not
significantly (p=0.5) (Figure S3). Similarly, L2-regularized logistic
regression, XGBoost, L2-regularized SVM with linear and radial basis
function kernel AUROC values were not significantly different from one
another and had median AUROC values of 0.680 {[}IQR 0.639-0.750{]},
0.679 {[}IQR 0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668
{[}IQR 0.639-0.750{]}, respectively. L1-regularized SVM with linear
kernel and decision tree had significantly lower AUROC values than the
other ML models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and
0.601 {[}IQR 0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly,
these results demonstrate that the most complex model (XGBoost) did not
have the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well as non-linear models.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
could indicate that the models were overfit to the training data. The
largest difference in median AUROCs was 0.021 in L1-regularized SVM with
linear kernel, followed by SVM with radial basis function kernel and
decision tree with a difference of 0.007 and 0.006, respectively
{[}Figure 2{]}. These differences were relatively small and gave us
confidence in our estimate of the generalization performance of the
models.

@@

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. The training times increased with
the complexity of the model and the number of potential hyperparameter
combinations. Also, the linear models trained faster than non-linear
models {[}Figures S1-S2; Figure 5{]}.

@@

We established a pipeline that can be generalized to any modeling method
that predicts a binary health outcome. We performed a random data-split
to create a training set (80\% of the data) and a held-out test set
(20\% of the data), which we used to evaluate predictive performance. We
used the AUROC metric to evaluate predictive performance as it is a
clinically relevant evaluation metric for our study. We repeated this
data-split 100 times to measure the possible variation in predictive
performance. During training, we tuned the model hyperparameters with a
repeated five-fold cross-validation. Despite the high number of features
microbiome datasets typically have, the models we built with this
pipeline generalized to the held-out test sets.

@@

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, (5)). The OTU abundance table was
generated by Sze et al (44), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (42, 45, 46);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1, and the
lowest would be 0.

@@

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter.For
L2-regularized logistic regression, L1 and L2-regularized SVM with
linear and radial basis function kernels, we tuned the cost
hyperparameter, which controls the regularization strength, where
smaller values specify stronger regularization. For SVM with radial
basis function kernel, we also tuned the sigma hyperparameter, which
determines the reach of a single training instance where, for a high
value of sigma, the SVM decision boundary will be dependent on the
points that are closest to the decision boundary. For the decision tree
model, we tuned the depth of the tree where the deeper the tree, the
more splits it has. For random forest, we tuned the number of features
to consider when looking for the best tree split. For XGBoost, we tuned
the learning rate and the fraction of samples used for fitting the
individual base learners. Performing a grid search for hyperparameter
selection might not be feasible for when there are more than two
hyperparameters to tune for. In such cases, it is more efficient to use
random search or recently developed tools such as Hyperband to identify
good hyperparameter configurations (47).

The computational burden during model training due to model complexity
was reduced by parallelizing segments of the ML pipeline. We
parallelized the training of each data-split. This allowed the 100
data-splits to be processed through the ML pipeline simultaneously at
the same time for each model. It is possible to further parallelize the
cross-validation step for each hyperparameter setting which we have not
performed in this study.

@@

\textbf{Acknowledgements.} We thank all the study participants of Great
Lakes-New England Early Detection Research Network. We would like to
thank the members of the Schloss lab for their valuable feedback. Salary
support for MR came from NIH grant 1R01CA215574. Salary support for PDS
came from NIH grants P30DK034933 and 1R01CA215574.

@@

@@

\textbf{Figure 3. Interpretation of the linear ML models.} The ranks of
absolute feature weights of (A) L1-regularized SVM with linear kernel,
(B) L2-regularized SVM with linear kernel, and (C) L2-regularized
logistic regression, were ranked from highest rank, 1, to lowest rank,
100, for each data-split. The feature ranks of the 20 highest ranked
OTUs based on their median ranks (median shown in black) are reported
here. OTUs that were associated with classifying a subject as being
healthy had negative signs and were shown in blue. OTUs that were
associated with classifying a subject having an SRN had positive signs
and were shown in red. \newpage
\includegraphics{Figure_4.png}

\textbf{Figure 5. Training times of seven ML models.} The median
training time was the highest for XGBoost and shortest for
L2-regularized logistic regression. \newpage

\captionof{table}{\textbf{Table S1.} An aspirational rubric for evaluating the rigor of ML practices applied to microbiome data.}
\small

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline

\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} (A) L2-regularized logistic
regression, (B) L1-regularized SVM with linear kernel, and (C)
L2-regularized SVM with linear kernel mean cross-validation AUROC values
when different hyperparameters were used in training the model. The
stars represent the highest performing hyperparameter setting for each
model. \newpage
\includegraphics[height=30cm, width=15cm]{Figure_S2.png}
